the rethinking way of doing the binary outcome modeling

### 2021-03-12 scripts/54_rethinking_binary.Rmd
it looks good then it is not

---
title: "the rethinking way of doing the binary outcome modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rethinking)
library(tidyverse)  
```

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(logdiv_s = scale(log(meta$simpson_reciprocal))) %>% 
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2))
```

```{r}
# visualize the var distribution 
table(meta$cr_d100)  
table(meta$toxicity)
hist(meta$simpson_reciprocal)
hist(log(meta$simpson_reciprocal)) # I think I should use log transformed
table(meta$center)
```
```{r}
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    logdivs = meta$logdiv_s
    )

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdivs + a[location] ,
        b ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE )
```

>>>>

---
title: "the rethinking way of doing the binary outcome modeling"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}
library(rethinking)
library(tidyverse)  
library(haven) #for reading sav data
library(sjstats) #for calculating intra-class correlation (ICC)
library(effects) #for plotting parameter effects
library(jtools) #for transformaing model summaries
library(ROCR)
```

# CR/Toxicity ~ alpha diversity + center

## Toxicity

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(div_std = scale(simpson_reciprocal)) %>%  # diversity is scaled 
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2)) %>%  # MSK 1 ; Upenn 2
  mutate(center = factor(center),
         toxicity = factor(toxicity, levels = c('no','yes')),
         cr_d100 = factor(cr_d100, levels = c('no','yes')))

meta %>% write_csv('../data/amplicon/stool/combined_2_meta_expanded.csv')
```

```{r}
# visualize the var distribution 
table(meta$cr_d100)  
table(meta$toxicity)
hist(meta$simpson_reciprocal)
hist(log(meta$simpson_reciprocal)) # I think I should use log transformed
table(meta$center)
table(meta$toxicity, meta$center)
table(meta$cr_d100, meta$center)   
range(meta$div_std)
```
```{r}
set.seed(123)
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    div_std = meta$div_std
    )

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*div_std + a[location] ,
        b ~ dnorm( 0 , 2), # the std of 2 is made by the later prior check
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

precis(mtox, depth = 2)
plot(precis(mtox, depth = 2) ) 
# there is baseline difference between penn and Msk and MSK has higher tox rage in general.
# it doesn't seem diversity has a clear association with tox rate after correcting for center wise difference
```

```{r}
# doing a prior check
prior <- extract.prior( mtox , n=1e4 ) 
# don't know how to put in logdivs into this prior check
divs_samp <- sample(meta$div_std, size = 1e4, replace = T)
p <- sapply( 1:2 , function(k) inv_logit( prior$b * divs_samp + prior$a[,k] ) )
mean( abs( p[,1] - p[,2] ) )
dens(abs( p[,1] - p[,2] ), adj = 0.1) # seems the difference is a bit too high
```

```{r}
# posterior prediction check

# the raw proportion of yes toxicity in each place
rawtox <- meta  %>% 
  count(center, toxicity) %>% 
  left_join(meta %>% 
              count(center, name = 'total'), by = 'center') %>% 
  group_by(center, toxicity) %>% 
  summarise(frac = round(n/total, 2)) %>% 
  filter(toxicity == 'yes')

# don't know how to do the post check since the diversity is continuous variable

```

```{r}
# I also want to do it in the glm way so that I can compare
lm_tox <- glm(formula = toxicity ~ simpson_reciprocal + center,
                    family = binomial(link = "logit"),
                    data = meta)
#summary(lm_tox)
summ(lm_tox, exp = T)# this is the log odds

```

## CR_d100

```{r}
mcr <- ulam( 
    alist( 
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- b*div_std + a[location] ,
        b ~ dnorm( 0 , 2), # the std of 2 is made by the later prior check
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

precis(mcr, depth = 2)
plot(precis(mcr, depth = 2) ) 
# there isn't much difference in baseline level between the centers
# diversity seems to be positively associating with CR

# if switching to standardized diversity, the positive association is less sure.

```

```{r}
# doing a prior check
prior <- extract.prior( mcr , n=1e4 ) 
# don't know how to put in logdivs into this prior check
divs_samp <- sample(meta$div_std, size = 1e4, replace = T)
p <- sapply( 1:2 , function(k) inv_logit( prior$b * divs_samp + prior$a[,k] ) )
mean( abs( p[,1] - p[,2] ) )
dens(abs( p[,1] - p[,2] ), adj = 0.1)
```

```{r}
# I also want to do it in the glm way so that I can compare
lm_cr <- glm(formula = cr_d100 ~ simpson_reciprocal + center,
                    family = binomial(link = "logit"),
                    data = meta)
#summary(lm_tox)
summ(lm_cr, exp = T)# this is the log odds
```
